From physical conditions to Hamiltonian and Lagrangian mechanics

### 2020-12-23 Articles/2020-ClassicalMechanicsPhil/Source/classical-phil-article.tex
Adding section on density/riemmanian geometry

>>>>

\subsection{Density and Riemmanian geometry}

It is helpful to take a quick look at how densities work in general so that we can better understand what is so special about them in the context in Hamiltonian mechanics. By density we mean the intuitive physical notion of a quantity divided by region it is found, like mass per volume or charge per area. Over a continuum, this is an integrable function $\rho(x)$. Mathematically, the concept is captured by the Radon–Nikodym derivative: we have a region $U$, we have two measures $m$ and $V$ and the derivative satistifies $m(U) = \int_U \rho dV$. This makes it clear that the units of $\rho$ depends on the units of both $m$ and $V$. More precisely, if $\rho(x, y)$ is a function of multiple variables, we will want $\int_U \rho(x, y) dx dy$ so that the units of $\rho$ are consistent with the units of the arguments. For example, in polar coordinates $\rho(r, \theta)$ will return mass over meter-radians.

This object is coordinate dependent because its definition requires us to make a choice of coordinates. Worse than that, it is not coordinate invariant: the density changes with the Jacobian. For example, changing polar to Cartesian we have $\rho(x,y) = |J| \rho(r, \theta) = r \rho(r, \theta)$. Since the Jacobian is a function of the point, this makes the value of the density, in general, not directly comparable at different points. If we have a uniform density in $x$ and $y$, we will have an $r$ dependence in $r$ and $\theta$. This dependence is introduced precisely because the area identified by $dr$ and $d\theta$ is $dr \,r d\theta$, which is a function of $r$.

We can look at this through the lens of differential geometry. First of all, note that in differential geometry there is also an object called density, but this is not the function $\rho$ we are looking for. A density in differential geometry is essentially a volume form (i.e. an alternating function from $n$ vectors to a real number) that is unaffected by changes in orientation. What we are looking for is really the component of a volume form $\nu = \rho(x^i) dx^1 \wedge dx^2 \wedge ... \wedge dx^n$. Note that $\nu$ combines the density and the volume, and that is why it is coordinate-free. Comparing the integrals we have $m(U) = \int \nu$ so essentially $\nu = dm$ which is the infinitesimal mass contribution we are integrating. But this means the value returned by $\nu$ does not depend only on how must stuff is there at the point, but on the size of the volume identified by the $n$ vectors given to the form.

In Riemannian geometry, we have a canonical volume form $dV = \epsilon \sqrt{|g_{ab}|} dx^1 \wedge dx^2 \wedge ... \wedge dx^n$ where $|g_{ab}|$ is the determinant of the metric tensor as expressed in local coordinates and $\epsilon$ is $1$ or $-1$ depending on the orientation. It can also be expressed as $\sqrt{|g_{ab}|}\epsilon_{a_1 ... a_n}$ in component notation where $\epsilon_{a_1 ... a_n}$ is the Levi-Civita symbol. We can then express any differential density $\nu = \hat{\rho} dV$ as the product of the volume density times a scalar function $\hat{\rho}$. By simply looking at the math, we may be mislead to think we were able to express the density in a coordinate independent way, but this is not the case. The quantity $|g_{ab}|$ depends on the choice of coordinates and is, in general, a function of the point. However, we can always find a set of coordinates that, at least locally, makes that quantity constant. The determinant $|g_{ab}|$ is the Jacobian of the transformation to those coordinates. For example, in Cartesian coordinates $|g_{ab}| = 1$ and the volume form $dV = dx \wedge dy$; in polar coordinates $|g_{ab}| = r^2$ and the volume form $dV = r dr \wedge d\theta$. This means that $\hat{\rho}$ is really expressing the density in a particularly convenient Cartesian-like coordinate system: one in which we can compare the density point-wise. We can also see that if we look at the units of $\hat{\rho}$, which is terms of inverse volume as expressed in the privileged system.\footnote{This is similar to what happens with the concept of proper time: it is defined in the comoving frame, so it is coordinate dependent. Yet, the expression $d\tau = \frac{1}{c} \sqrt{g_{ab} dx^a dx^b}$ is coordinate invariant.}

We can now look at what happens for distributions over position and velocity. Note that the velocities transform in the same way (i.e. controvariantly) than differentials in position. Therefore $\rho(x^i, v^j) = \sqrt{|g_{ab}|} \sqrt{|g_{ab}|} \rho(\hat{x}^i, \hat{v}^j) = |g_{ab}| \rho(\hat{x}^i, \hat{v}^j)$ $dx^a$. The determinant of the metric tensor is therefore directly the volume of the tangent bundle. The relationship to the density as expressed in phase space is therefore the following:
\begin{equation}
	\rho(q^i, p_j) = |g_{ab}| \rho(x^i, v^j).
\end{equation}
A change of coordinates will induce a controvariant change in the velocity and a covariant change in the momentum. The left side will remain invariant. The two terms on the right size will change in the opposite way so that their product is invariant. Focusing on densities, then, brings about a physically motivated way to understand the relationship between the cotangent and the tangent bundle, where the metric tensor plays a role in defining the geometry of the second in a similar way that the symplectic form defines the geometry of the first: it defines the count of states so that densities can be properly tracked.
